---
title: "Managing multiple data-frames"
subtitle: "Presented to EdinbR R-Users Group, 2018-07-18"
author: "Russ Hyde, University of Glasgow"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    df_print: paged
---

<!-- 
PLAN:
- data-modelling
    - tidy data
    - 2.5th normal form
    - tidying-up of a simple data-frame into two conceptually-linked data-frames
    - ExpressionSet/Matrix as an untidy example
- tidygraph
    - tidy data version of igraph
    - examples of using tidygraph
    - the activate verb
- polyply
-->

<!--
  Styling to allow footnotes to appear on the same slide as their definition
  - Borrowed from https://stackoverflow.com/questions/42690955
-->

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
</style>

<script
  src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"
  ></script>

<script>
  $(document).ready(function() {
    $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

    $('footnote').each(function(index) {
      var text  = $(this).html();
      var fnNum = (index+1).toString().sup();
      $(this).html(text + fnNum);

      var footnote   = fnNum + ': ' + $(this).attr('content') + '<br/>';
      var oldContent = $(this).parents('slide').children('div.footnotes').html();
      var newContent = oldContent + footnote;
      $(this).parents('slide').children('div.footnotes').html(newContent);
    });
  });
</script>

<!-- Packages -->

```{r, echo = FALSE}
suppressPackageStartupMessages({
  library(Biobase)
  library(DiagrammeR)
  library(igraph)
  library(igraphdata)
  library(pryr)
  library(tibble)
  library(tidyr)
  library(tidygraph)
})
```

<!-- Presentation code -->

## Social:

- @POG_LRC
    - The Paul O'Gorman Leukaemia Research Centre

- @haematobot
    - Personal mumblings about code / analysis / bioinformatics and seemingly
    very little else ...

- https://biolearnr.blogspot.com/
    - Even more mumblings

# Data-Modelling

<!-- Is tidy-data always the answer? -->

## Tidy Data and the Normal-Forms {.build}

In <footnote content="http://vita.had.co.nz/papers/tidy-data.html">tidy data</footnote>:

- TD1 - Each variable forms a column.
- TD2 - Each observation forms a row.
- TD3 - Each type of observational unit forms a table.
- [TD4 - A key permitting table-joins is present]

See also, Boyce-Codd
<footnote content="https://en.wikipedia.org/wiki/First_normal_form">Normal-Forms</footnote> and relational-database-design.

## Common _Untidy_ Data Structures

Tidy-data / normal-forms help reduce duplication and play nicely with `ggplot2`

But untidy data-structures are useful if they:

- increase access efficiency

- decrease code complexity

- play nicely with other important things

## `Biobase::ExpressionSet`

```{r, message=FALSE, eval=FALSE}
Biobase::ExpressionSet()
```

<!-- TODO: colour the sides of the nodes according to matching dimensions
     - eg, colour of assayData and featureData verticals should match,
           colour of phenoData vertical should match assayData horizontal
           -->

```{r, echo = FALSE}
grViz("
digraph eset_boxes {

  # graph, node, and edge definitions
  graph [compound = true, nodesep = .5, ranksep = .25,
         color = crimson]

  node [fontname = Helvetica, fontcolor = darkslategray,
        shape = rectangle, fixedsize = true, width = 3,
        color = darkslategray]

  edge [color = grey, arrowhead = none, arrowtail = none]

  subgraph eset {
    # nodes
    node [fixedsize = true]
    A [label = 'assayData\n(matrix)\n\nnrow=|genes|\nncol=|samples|', height = 3, width = 2]
    B [label = 'featureData\n(data-frame)\n\nnrow=|genes|', height = 3, width = 2.5]
    C [label = 'phenoData\n(data-frame)\n\nnrow=|samples|', height = 2, width = 2.5]
  
    # edges
    A->B A->C
  }

  ExpressionSet         [width = 2]
  ExpressionSet -> A    [lhead = eset]
  ExpressionSet -> protocolData
  ExpressionSet -> experimentData
  ExpressionSet -> Annotation
}
")
```

Figure made with `DiagrammeR`

## `Biobase::ExpressionSet` (cont.)

Conversion of the `assayData` to meet tidy-data standards:

```{r, echo = FALSE}
m <- as.matrix(data.frame(
  sample1 = c(12.2, 19.1, 0.5),
  sample2 = c(111, 10.5, 3.4),
  sample3 = c(129, 123, 1.1),
  row.names = paste0("gene", 1:3)
))
```

```{r}
m # our assayData
```

Doesn't meet tidy-data standards:

- rows correspond to features, columns to samples
- not all variables are in columns (since row-IDs are meaningful)
- entries are the same 'type' of variable

----

Easy <footnote content="... or as.data.frame / rownames_to_column / gather">fix</footnote>:

```{r}
m2 <- reshape2::melt(
    m,
    varnames = c("feature_id", "sample_id"),
    as.is = TRUE
  )

head(m2, 4)
```

## But ...

- Matrix representation was more dense

- Lost all encapsulation

- (After modifying featureData / phenoData to match)
    - Have to join rather than index
    - Have to keep track of multiple data-frames, rather than one
      data-structure

<!-- Remove this next section: -->

## That multi-data-frame _thing_

For a reasonably complex project:

- tidy-data / normal-forms mean more data-frames

Wanted:

- a lightweight approach to working with multiple 'conceptually-related'
data-frames

- that plays nicely with `tidyverse` verbs

- that feeds into `ggplot2`

- that plays nicely with untidy data-structures I use _all the time_

# `tidygraph` already does this

## Basics of 'graph theory' speak {.columns-2}

```{r, echo = FALSE}
data(Koenigsberg, package="igraphdata")
par(mfrow=c(1, 3))
plot(Koenigsberg)
```

A graph is made up of two sets:

- _V_, a set of vertices:
    - aka nodes, actors, ...

- _E_, a set of edges:
    - pairwise relationships between vertices
    - aka interactions, lines, arcs, ...

- Need to store attributes for both nodes and edges

## `tbl_graph` data structure

```{r}
data("Koenigsberg", package = "igraphdata")
tg <- tidygraph::as_tbl_graph(Koenigsberg)

# Nodes data shows up first:
tg
```

## `tbl_graph` data structure

```{r}
# If we make the 'edges' active, the edge-data shows up first:
activate(tg, edges)
```

## The `activate` verb

Think of the `tbl_graph` as `list[nodes, edges]`

To modify the contents of a given data-frame, `activate` it:

```{r}
tg %>%
  activate(edges) %>%
  mutate(weight = nchar(name))
```

# `polyply` and multiple, linked data-frames

## `polyply` {.build}

`tidygraph` is really a wrapper around the package `igraph`

- `igraph` stores graphs in a slightly different format

But I liked the idea of holding connected data-frames in one place

## (WORK-IN-PROGRESS: so please be nice)

