---
title: "Using `polyply`"
author:
- name: "Russ Hyde"
  affiliation: "University of Glasgow"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        df_print: kable
vignette: >
  %\VignetteIndexEntry{using_polyply}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Import the Package

`polyply` was written to work with the `tidyverse` / `dplyr` approach to
manipulating data-frames in `R`. Load in the packages for the current vignette:

```{r}
suppressPackageStartupMessages({
  library(polyply)

  library(dplyr)
  library(purrr)

  requireNamespace("ggplot2", quietly = TRUE)
})
```

# PURPOSE!

<!-- TODO -->

# ANIMALS!

`polyply` contains a few datasets. The datasets `animals`, `common_to_species`
and `taxonomy` are logically linked. They were either obtained from (`animals`
was from `MASS`) or using R packages. See the vignette
`origins_of_the_datasets` for more details.

`animals` contains the brain and body weight of a few animals. It has been
tidied-up relative to the `MASS` version by pulling the common-name of the
animals into the columns of the data-frame (they were previously stored in the
rownames).

```{r}
data(animals, package = "polyply")
head(animals)
dim(animals)
```

`common_to_species` contains the common-name of the animals and the
species-name (that is, the genus / species pair or taxonomic name) for each
animal. A formal species-name couldn't be found for one of the animals (nor
could any other reference to that animal).

```{r}
data(common_to_species, package = "polyply")
head(common_to_species)
dim(common_to_species)
filter(common_to_species, is.na(species))
```

Finally, `taxonomy` contains various taxonomic information (the 'family' and
'order') about each species. See [ Wikipedia ](
https://en.wikipedia.org/wiki/Taxonomy_(biology)
) for more details about taxonomy. Again, there is some missing data in this
data-frame.

```{r}
data(taxonomy, package = "polyply")
head(taxonomy)
filter(taxonomy, is.na(family) | is.na(order))
```

## Brain-size and Body-mass

Large brains and large bodies go together. To show this, we are going to make a
few charts using the `animals` dataset, and use the associated taxonomic
information to annotate the charts.

First we make a little helper function to reduce duplication in the plotting
calls.

```{r}
gg_helper <- function(df) {
  ggplot2::ggplot(
    df,
    ggplot2::aes(x = body, y = brain)
  ) +
    ggplot2::geom_point() +
    ggplot2::scale_x_log10(
      name = "Body Weight / kg",
      breaks = c(0.01, 1, 100, 10000),
      labels = c("0.01", "1", "100", "10,000"),
      limits = c(0.01, NA)
    ) +
    ggplot2::scale_y_log10(
      name = "Brain Weight / g",
      breaks = c(0.1, 10, 1000),
      labels = c("0.1", "10", "1,000"),
      limits = c(0.1, NA)
    )
}
```

```{r, fig.width = 6}
animals %>%
  gg_helper() +
  ggplot2::ggtitle(
    label = "`ggplot2` is great isn't it?",
    subtitle = "Try `ggrepel` if you want to get all the text-labels visible"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = common_name),
    position = ggplot2::position_jitter(0.2, 0.1),
    col = colors()[123],
    size = 3,
    check_overlap = TRUE
  )
```

That's a perfectly nice plot. But it's a bit odd

- we've got dinosaurs and mammals on the same chart (unsurprisingly, dinosaurs
  are something of an outgroup - what with them not being mammals - maybe a
  broader set of vertebrate classes should have been presented)

- the mammals are of varying different types (rodents, carnivores, primates)

- there's five orders of magnitude between the different body weights

- some of the body weights are measurable, and some of them have
been inferred....

Let's make a different chart, where we separate the species up based on their
taxonomic order (rodent, carnivore ...).

### The `dplyr` route:

To make the required chart, we have to associate each row of `animals` with the
corresponding `order` in the `taxonomy` data-frame. So we have to join the
three data-frames together. The workflow for doing this:

1. Combine data-frames together

1. Call `ggplot`, set any plot-aesthetics and add any required geometry

1. Format the chart

```{r, fig.width = 6}
# Combine the data-frames together
animals %>%
  left_join(common_to_species) %>%
  left_join(taxonomy) %>%
  # Make the chart
  gg_helper() +
  # Add some extra formatting tweaks
  ggplot2::geom_point(ggplot2::aes(col = order))
```

There's a few different ways to do that first data-frame combining step. Here
we used `dplyr`'s `left_join` function. Since we only used `left_join`, we
could have used the following, more 'functional programming' approach to do the
same thing:

~~~
// This ...
purrr::reduce(
  list(animals, common_to_species, taxonomy),
  left_join
)

// is equivalent to ...
left_join(
  left_join(
    animals, common_to_species
  ),
  taxonomy
)
~~~

### (One of) The `polyply` way(s)

```{r, fig.width = 6}
poly_frame(
  animals, common_to_species, taxonomy,
  merge_fn = function(x) reduce(x, left_join)
) %>%
  merge() %>%
  gg_helper() +
  ggplot2::geom_point(ggplot2::aes(col = order))
```

----

END OF VIGNETTE

----

Vignettes are long form documentation commonly included in packages. Because
they are part of the distribution of the package, they need to be as compact as
possible. The `html_vignette` output type provides a custom style sheet (and
tweaks some options) to ensure that the resulting html is as small as possible.
The `html_vignette` format:

- Never uses retina figures

- Has a smaller default figure size

- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block
above. These are required in order to instruct R how to build the vignette.
Note that you should change the `title` field and the `\VignetteIndexEntry` to
match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme
you can specify your own CSS in the document metadata as follows:

    output:
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images
side-by-side.

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in
**knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$,
footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
